\mychapter{12}{Lesson 12} %181107
\section{Hashing}

Remember one solution to domain-exetension for \prf{}s, as a composition of a \prf{} $F$ with an almost universal hash function $H$. Hash functions compress their arguments to some ``fingerprint'', which is asssumed to be unique. However, since this compression in this context inherently introduces information loss, it is not guaranteed that every message gets its own unique fingerprint; indeed, there will be some instances where two messages yield the same hash value, or in other words, the hashes \emph{collide}. It is desirable for a hash function to be \emph{resistant} to these events, meaning that it is hard to reproduce such collisions.

\begin{definition}
    A hash function family $H$ is deemed \emph{collision-resistant}, denoted as $H \in \crh$ iff the probability of finding a collision is negligible, even when knowing the seed $s$. Formally:
    \[
        \forall \adversary \in \ppt \implies \Pr(\cryptog{crh}(\lambda) = 1) \in \negl(\lambda)
    \]

    \begin{cryptogame}
        {crh}
        {The \emph{collision-resistance} game}
        {crh}

        \receive{$s \pickUAR \binary^\lambda$}{$s$}{}

        \send{}{$x, y$}{\textsc{Output 1 iff} $h_s(x) = h_s(y)$}
        
    \end{cryptogame}
\end{definition}

A note: before, we were dealing with unbounded adversaries, and the key was hidden. Now the key is public, but the adversary must be efficient.

\begin{exercise}
    Let $\Pi$ be a \ufcma{} authentication scheme over the message space $\binary^n$. Show that $\Pi' = (\Tag', \Ver') : \Tag'_{k, s}(m) = \Tag_k(h_s(m))$ is \ufcma-secure over $\binary^l$, where $l \in \poly(n)$, as long as $h_s$ itself is \crh.
\end{exercise}

\subsection{Merkle-Damg\r{a}rd construction}

A construct for starters has been defined by Ralph Merkle and Ivan Damg\r{a}rd, which revolves around the use of a \crh{} function $h_s \in \binary^{n + 1} \to \binary^n$. For now, consider the messages to be of an arbitrarily fixed length $l$. The steps to follow are:

\begin{enumerate}
    \item Let $\textsc{iv} \in \binary^b$ be an \emph{initialization vector}
    \item Initialize $t_0 := \textsc{iv}$
    \item For each bit $b_i$ of the message to hash $m$:
    \begin{enumerate}
        \item Compute $t_i := h_s(b_i, t_{i - 1})$
    \end{enumerate}
    \item Return $\phi = t_n$
\end{enumerate}

% AP190203: To review, it is somewhat confused
%First step: Compress the original message (assuming fixed size) by one bit

%Let $H$ be a one-bit shrinking function. Then, it can be used to construct a hash function H' that splits an arbitrary-size message into fixed-size blocks, apply H onto them, and return a digest of fixed length. This is exemplified by the diagram in figure \ref{fig:mdbase}

\begin{figure}
    \centering

    \tikzstyle{int}   = [draw, minimum size=2em]
    \tikzstyle{empty} = [minimum size=2em]
    \tikzstyle{init}  = [pin edge={to-,thin,black}]

    \begin{tikzpicture}[node distance = 1.9cm, auto, >=latex']

        \node (a) [empty] {$\textsc{iv}$};
        \node (r) [int, pin={[init]above:$b_1$}] [right of=a] {$h_s$};
        \node (d) [int, pin={[init]above:$b_2$}] [right of=r] {$h_s$};
        \node (e) [int, pin={[init]above:$b_3$}] [right of=d] {$h_s$};
        \node (f) [empty] [right of=e] {$...$};
        \node (g) [int, pin={[init]above:$b_n$}] [right of=f] {$h_s$};
        \node (h) [empty, right of=g] {$\phi$};

        \path[->] (a) edge (r);
        \path[->] (r) edge node {$t_1$} (d);
        \path[->] (d) edge node {$t_2$} (e);
        \path[->] (e) edge node {$t_3$} (f);
        \path[->] (f) edge node {$t_{n-1}$} (g);
        \path[->] (g) edge (h);
    
    \end{tikzpicture}
    \caption{Basic outline of a Merkle-Damg\r{a}rd construction}
    \label{fig:mdbase}
\end{figure}

Figure \ref{fig:mdbase} depicts a general view of the algorithm. Let it be denoted as another hash function $h'_s \in \binary^l \to \binary^n$.

\begin{theorem}
    The construction $H'$ obtained by Merkle-Damg\r{a}rd is a \textsc{crh} function.
\end{theorem}

% AP190204: Tentative
\begin{proof}
    Assume $H'$ can be broken efficiently by a distinguisher $\distinguisher^{\crh}$, meaning that finding two distinct block sequences that give the same hash is easy. Consider the reduction to $H$'s \crh-ness in figure \ref{cryptoredux:mdcrh}

    \begin{cryptoredux}
        {mdcrh}
        {Breaking the underlying \crh-ness of $H$}
        {h-crh}
        {h'-crh}[3]

        \receive{}{$s$}{}
        \invoke{}{$s$}{}
        \return{}{$x, y$}{}

        \cseqdelay

        % AP190911: need a better way to convey meaning here
        \send{\shortstack[r]{
            \textsc{Find} $j :$ \\
            $h'_{s, (j - 1)}(x) \neq h'_{s, (j - 1)}(y)$ \\
            $\wedge\: h'_{s, (j)}(x) = h'_{s, (j)}(y)$
        }}{$h'_{s, (j - 1)}(x), h'_{s, (j - 1)}(y)$}{}

    \end{cryptoredux}

    Ignore same blocks: find largest j such that:
    \[
        (b_j, x_{j - 1}) \neq (b'_j, y_{j - 1}) \wedge h_s(b_j, x_{j - 1}) = h_s(b'_j, y_{j - 1})
    \]
    this implies the rest of th message is equal, then the resulting final hash will be equal, thus for $j > 0$ we have a collision.

\end{proof}

%AP190911: Not clear at all
\subsubsection{Domain extension}

In order to adapt the MD-construct to messages of variable length, some sort of ``strengthening'' is required: 

% Also called MD-strengthening
\begin{lemma}[Length padding]
    Let $H_s \in \binary^{n + l} \to \binary^n$, then:
    %AP190911: This needs to be rewritten correctly
    \[
        H'_s = H_s(\left<l'\right>, H_s(x_{l'}, \dots , H_s(x_1, 0^n) \dots)), |l'|, |x_i| \in \binary^c 
    \]
\end{lemma}

This is an example of padding which encodes the message length in itself.

\begin{theorem}
    The strengthened construct is collision resistant for variable-length messages.
\end{theorem}

\begin{proof}
    Hint: similar as above, case by case
\end{proof}

\subsubsection{Merkle trees}

---

\subsection{Compression functions}

Let ($\textsf{Gen} : 0 \mapsto (\pk, \sk), f, g)$ be a \pke{} scheme, where the functions $f$ and $g$ are keyed \prp{}s. A \emph{claw} is a couple of values $(x, x')$ such that:
\[
    f(\pk, x) = g(\pk, x')
\]

\begin{cryptogame}
    {cfp}
    {The game of claw-free permutations}
    {cfp}
    
    \receive{$(\pk, \sk) \pickUAR \textsf{Gen}$}{$\pk$}{}
    \send{}{$x, x'$}{\textsc{Output 1 iff} $f(\pk, x) = g(\pk, x')$}

\end{cryptogame}

\begin{theorem}
    Assuming $\mathcal{F}$ is claw-free, then $h_\pk$ is \crh{} from $n + l$ bits to $n$.
\end{theorem}

\subsubsection{Davies-Meyer construct}

% AP190911: How does AES come in here?!
\begin{definition}
    $x_{i + 1} = E_k(x_i) \xor x_i$, maps n+$\lambda$ to n. E is AES
\end{definition}
